---
title: intro
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{intro}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---


# GSA_agemodeling

Welcome to the GSA Short Course on *Age-Depth Modeling of Sedimentary Deposits*. This course had four sessions:

1. A prep check-in on Sep 13, to make sure all relevant software on your computer is working
2. On Sep 15 we discussed radiocarbon dating and calibration, and introduce how to make simple age-models
3. On Sep 22 we looked at the theory and implementation of Bayesian age-depth models (mainly `rbacon`)
4. Then finally on Sep 29 we looked at details of rbacon, introduced rplum Pb-210 age-modelling and looked at some age-models for cores from the participants.

All sessions took place at 9-11am PDT (Pacific Daylight Time) on Zoom.

You will need to have a recent version of [R](http://r-project.org) installed for your operating system, at least version *4.1.1*. If you want, you can also install [RStudio](http://www.rstudio.com). Note that the version numbering differs between `R` and `Rstudio`; most important is to have a recent version of `R` installed.
Please install the following R packages, by typing or pasting the below command into the R terminal:

```{r, eval=FALSE}
install.packages(c("IntCal", "clam", "rbacon", "coffee", "rplum"))
```

After these packages are installed, please try loading both and running `clam` and `rbacon`, so that we can be sure everything will work as expected during the sessions:

```{r, eval=FALSE}
library(clam)
clam()
library(rbacon)
Bacon()
```



[next: session 1](session_1.html)

[session 1](session_1.html)<br>
[session 2a](session_2a.html)<br>
[session 2b](session_2b.html)<br>
[session 3](session_3.html)

